Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-15 manuscript.tex
initial rewrite of developmental

\documentclass[10pt]{article}
\usepackage[top=20mm, bottom=20mm, left=20mm, right=20mm]{geometry}
\geometry{a4paper}
\geometry{portrait}
\usepackage[utf8]{inputenc}	% Comment out for xelatex; must use with pdflatex
\usepackage[T1]{fontenc}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{fontspec}
\usepackage{epstopdf}
\usepackage{float}
\usepackage[hang]{footmisc}
\usepackage{endnotes}
\usepackage{listingsutf8}
%\restylefloat{table}
\usepackage{xcolor}
\usepackage{colortbl}
%\usepackage{booktabs}
\usepackage{tabu}
\usepackage{caption}
\usepackage{textcomp}
\usepackage{textgreek}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\usepackage{authblk}
\usepackage[numbers]{natbib}
\bibliographystyle{plos2015}

@@

To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours. In general, higher number of components and neighbours has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (Fig. S1). Running MCMC algorithm recovers the performances of the majority of the configurations (Fig. S2).

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
\end{figure}

Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig. \ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$). In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3. Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them. It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig. \ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.

These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter. These data also hint at the possibility to identify rare cell types in larger populations.

@@

When grouping is performed by optimization of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package SquidPy \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors. At default resolution, modularity optimization resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig. \ref{Figure_Visium}). When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.

@@

\textcolor{red}{TO BE UPDATED, possibly with Visium data. The point here should be that hierarchy is more consistent with tissue type, then we talk about development}
The hierarchical model of cell groups implies that a relationship exists between groups. We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation. To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig. S3A). After applying nSBM, we identify 27 groups at level 3 of the hierarchy (Fig. S3B), compared to the 24 using Leiden method at default resolution (Fig. S4). We found that the hierarchy proposed by our model is consistent with the developmental model (Fig. \ref{Figure2}). Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (Fig. S4); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.

>>>>

\documentclass[10pt]{article}
\usepackage[top=20mm, bottom=20mm, left=20mm, right=20mm]{geometry}
\geometry{a4paper}
\geometry{portrait}
\usepackage[utf8]{inputenc}	% Comment out for xelatex; must use with pdflatex
\usepackage[T1]{fontenc}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{fontspec}
\usepackage{epstopdf}
\usepackage{float}
\usepackage[hang]{footmisc}
\usepackage{endnotes}
\usepackage{listingsutf8}
%\restylefloat{table}
\usepackage{xcolor}
\usepackage{colortbl}
%\usepackage{booktabs}
\usepackage{tabu}
\usepackage{caption}
\usepackage{textcomp}
\usepackage{textgreek}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\usepackage{authblk}
\usepackage[numbers]{natbib}
\usepackage{xr}
\externaldocument{supplementary}

\bibliographystyle{plos2015}

@@

%To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours. In general, higher number of components and neighbours has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (Fig. S1). Running MCMC algorithm recovers the performances of the majority of the configurations (Fig. S2).

%\begin{figure}[H]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
%\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
%\end{figure}

%Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig. \ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$). In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3. Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them. It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
%As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig. \ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.

%These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter. These data also hint at the possibility to identify rare cell types in larger populations.

@@

When grouping is performed by optimization of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package SquidPy \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors. At default resolution, modularity optimization resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig. \ref{Figure_Visium}). When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer, suggesting that the standard approach may introduce biases.

@@

In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameters. We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig. \textcolor{red}{S3A}). After applying Leiden method at default resolution we identified 24 groups. By lowering the $\gamma$ parameter we could observe cell groups that merge and split at different resolutions in a non hierarchcal manner (Fig. \ref{Figure_Paul15_Leiden_r}).

### 2021-04-15 supplementary.tex
initial rewrite of developmental

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S4.png}
\caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1. Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution \emph{r}=0.3 splits in groups 2 and 0 at resolution \emph{r}=0.2.}\label{MPFigure:76168960-F76F-44B1-A5FA-A31681CDAFFD}
\end{figure}

>>>>

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Paul15_Leiden_r.png}
\caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1. Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ } in the Erythroid branch groups 7 and 15 at resolution \emph{r}=1 are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
\end{figure}
